---
title: "Re_Analysis"
author: "Ugochukwu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Introduction
 

# load required libraries / additional files

```{r Library,message=FALSE, echo=FALSE }

if(!require(openxlsx)){
  install.packages("openxlsx", repos = "https://www.stats.bris.ac.uk/R/")
  library(openxlsx)}

if(!require(readxl)){
  install.packages("readxl", repos = "https://www.stats.bris.ac.uk/R/")
  library(readxl)}

if(!require(readr)){
  install.packages("readr", repos = "https://www.stats.bris.ac.uk/R/")
  library(readr)}

if(!require(doParallel)){
    install.packages("doParallel", repos = "https://www.stats.bris.ac.uk/R/")
    library(doParallel)}

if(!require(vip)){
    install.packages("vip", repos = "https://www.stats.bris.ac.uk/R/")
    library(vip)}

if(!require(doParallel)){
    install.packages("doParallel", repos = "https://www.stats.bris.ac.uk/R/")
    library(doParallel)}
 
if(!require(Hmisc)){
  install.packages("Hmisc", repos = "https://www.stats.bris.ac.uk/R/")
  library(Hmisc)}

if(!require(corrplot)){
  install.packages("corrplot", repos = "https://www.stats.bris.ac.uk/R/")
  library(corrplot)}

if(!require(dplyr)){
  install.packages("dplyr", repos = "https://www.stats.bris.ac.uk/R/")
  library(dplyr)}

if(!require(ggplot2)){
  install.packages("ggplot2", repos = "https://www.stats.bris.ac.uk/R/")
  library(ggplot2)}

if(!require(knitr)){
  install.packages("knitr",repos = "https://www.stats.bris.ac.uk/R/")
  library(knitr)}

if(!require(class)){
  install.packages("class",repos = "https://www.stats.bris.ac.uk/R/")
  library(class)}

if(!require(lme4)){
  install.packages("lme4", repos = "https://www.stats.bris.ac.uk/R/")
  library(lme4)}

if(!require(MASS)){
  install.packages("MASS", repos = "https://www.stats.bris.ac.uk/R/")
  library(MASS)}

if(!require(tidyverse)){
  install.packages("tidyverse", repos = "https://www.stats.bris.ac.uk/R/")
  library(tidyverse)}

if(!require(assertr)){
  install.packages("assertr", repos = "https://www.stats.bris.ac.uk/R/")
  library(assertr)}

if(!require(ggpubr)){
  install.packages("ggpubr", repos = "https://www.stats.bris.ac.uk/R/")
  library(ggpubr)}

if(!require(lubridate)){
  install.packages("lubridate", repos = "https://www.stats.bris.ac.uk/R/")
  library(lubridate)}

if(!require(flextable)){
  install.packages("flextable", repos = "https://www.stats.bris.ac.uk/R/")
  library(flextable)}

if(!require(patchwork)){
  install.packages("patchwork", repos = "https://www.stats.bris.ac.uk/R/")
  library(patchwork)}

if(!require(randomForest)){
  install.packages("randomForest", repos = "https://www.stats.bris.ac.uk/R/")
  library(randomForest)}

if(!require(pROC)){
  install.packages("pROC", repos = "https://www.stats.bris.ac.uk/R/")
  library(pROC)}

if(!require(officer)){
  install.packages("officer", repos = "https://www.stats.bris.ac.uk/R/")
  library(officer)} 

```


```{r Recode Responce Variable for Binary Classification, include=FALSE}

if(!require(caret)){
  install.packages("caret", repos = "https://www.stats.bris.ac.uk/R/")
  library(caret)}
```



***START   DATA PREP***
```{r Load the dataset, echo=FALSE}

#setwd("C:/Users/sixsix - /Users/sixsix/Desktop/MY R CODES AND FILE") #  **Make sure you set your working Dir**

df <- read.csv("diabetic_data.csv")

#View(df)
dim(df)
sum(is.na(df))
```


```{r}
#Binary Classification – With  focus on early readmissions (within 30 days)

## readmitted = 1 if readmitted == '<30 Days' and 0 if readmitted == '>30 Days' or readmitted == 'NO'

df <- df %>% mutate(readmitted = ifelse(readmitted == "<30", 1, 0))

#Check sum and % of missing per column

sum(is.na(df))
```


```{r}
# Calculate and format proportion of missing values per column
na_summary <- sapply(df, function(x) sum(is.na(x)) / length(x)) %>%
  round(2) %>%
  sort(decreasing = TRUE) %>%
  as.data.frame()

# Format data frame for flextable
na_summary <- tibble::rownames_to_column(na_summary, var = "Factors")
colnames(na_summary)[2] <- "Proportion_Missing"

# Display using flextable
flextable(na_summary)
```



```{r}
# Drop high-missing columns
df = subset(df , select = -c(`weight`,`payer_code`,`medical_specialty`))
df

```



```{r}
# Simple mode function
mode_fun <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Correct imputation
df_imputed <- df %>%
  mutate(
    across(where(is.numeric),
           ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)), #mean for the numeric
    across(where(~ is.factor(.) || is.character(.)),
           ~ ifelse(is.na(.), mode_fun(.), .)) #mode for the categorical
  )

sum(is.na(df_imputed))  # should be 0

df_imputed

df <- df_imputed %>%
  distinct() %>%   # now truly removes duplicates
  subset(select = -c(`encounter_id`,`patient_nbr`,`examide`,`citoglipton`))

dim(df) # 7 redundant variables have been removed.
```


```{r}
#convert the following to factors

df$race <- as.factor(df$race)
df$gender <- as.factor(df$gender)
df$age <- as.factor(df$age)
df$readmitted <- as.factor(df$readmitted)
```


```{r}
##Convert and Recode Lab Test Results to ordered factors 

df$max_glu_serum <- factor(df$max_glu_serum, levels = c("None", "Norm", ">200", ">300"), ordered = TRUE)
df$A1Cresult <- factor(df$A1Cresult, levels = c("None", "Norm", ">7", ">8"), ordered = TRUE)
```


```{r}
##Convert Medication Use to Factors

med_columns <- names(df)[grep("metformin|insulin|glipizide|glyburide|pioglitazone|rosiglitazone|change|diabetesMed|examide|citoglipton|tolazamide|troglitazone|miglitol|examide|acarbose|tolbutamide|glimepiride|chlorpropamide|nateglinide|repaglinide", names(df))]

df[med_columns] <- lapply(df[med_columns], factor)
str(df)
```


#**Exploratory Data Analysis (EDA)**

```{r}
# Binary target
ggplot(df, aes(x = factor(readmitted), fill = factor(readmitted))) +
  
  geom_bar() + 
  
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.2) +
  
  scale_fill_manual(values = c("0" = "lightgreen", "1" = "Red"), labels = c("No", "Within 30 days")) +
  
  labs(title = "Readmission Distribution", x = "Readmission (Binary)", y = "Count")
```





#**Numerical Feature Summary**

```{r}
# Select numeric columns

numeric_cols <- names(df)[sapply(df, is.numeric)]

numeric_cols

```

```{r,Numerical Feature Summary,echo=TRUE}
# Histograms for numeric variables
df %>%
  dplyr::select(all_of(numeric_cols)) %>%  # use all_of() for safety
  
  gather(key = "Feature", value = "Value") %>%
  
  ggplot(aes(x = Value)) +
  
  geom_histogram(fill = "lightgreen", bins = 30) +
  
  facet_wrap(~Feature, scales = "free") +
  
  theme_minimal()
```

#t-test/ANOVA for Numerical Features

```{r,t-test / ANOVA for Numerical Features, echo=TRUE}
t_results <- lapply(numeric_cols, function(col) {
  
  test <- t.test(df[[col]] ~ df$readmitted)
  
  return(data.frame(Feature = col, p_value = test$p.value))
  
}) %>% bind_rows()

# Sort by significance
t_results <- t_results %>% arrange(p_value)

print(t_results)

sig_nums <- t_results %>% filter(p_value < 0.05)
```

# Plot only significant numeric features

```{r,Boxplots for Significant Numeric Features, echo=T}
for (feature in sig_nums$Feature) {
  p <- ggplot(df, aes_string(x = factor(df$readmitted), y = feature, fill = factor(df$readmitted))) +
    geom_boxplot() +
    scale_fill_manual(values = c("0" = "skyblue", "1" = "tomato"),
                      labels = c("No", "Within 30 days")) +
    labs(title = paste("Distribution of", feature, "by Readmission"),
         x = "Readmission (Binary)", y = feature) +
    theme_minimal()
  print(p)
}
df %>%
  
  dplyr::select(all_of(sig_nums$Feature), readmitted) %>%
  
  gather(key = "Feature", value = "Value", -readmitted) %>%
  
  ggplot(aes(x = factor(readmitted), y = Value, fill = factor(readmitted))) +
  
  geom_boxplot() +
  
  facet_wrap(~Feature, scales = "free", ncol = 3) +
  
  scale_fill_manual(values = c("0" = "skyblue", "1" = "tomato"),
                    
                    labels = c("No", "Within 30 days")) +
  
  labs(x = "Readmission (Binary)", y = "Value",
       
       title = "Numeric Feature Distributions by Readmission Status") +
  
  theme_minimal()

```

#Categorical Feature Summary

```{r,Categorical Feature Summary, echo=FALSE, echo=TRUE}

# Select categorical/ factor variables
factor_cols <- names(df)[sapply(df, is.factor)]
factor_cols
```


```{r,Bar plots for categorical variables, echo=TRUE}
#Categorical Feature Summary
df %>%
  dplyr::select(all_of(factor_cols)) %>%       # select categorical features
  mutate(across(everything(), as.character)) %>% # ensure all are characters
  pivot_longer(cols = everything(), 
               names_to = "Feature", 
               values_to = "Value") %>%        # long format for ggplot
  ggplot(aes(x = Value, fill = Value)) +       # plot distribution
  geom_bar() +
  facet_wrap(~Feature, scales = "free") +      # facet by feature
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#Test of Association

```{r,Chi-Square Test for Categorical Features, echo=TRUE}
# Run chi-square tests (Binary target recommended)
chi_results <- lapply(factor_cols, function(col) {
  
  tbl <- table(df[[col]], df$readmitted)
  
  # Only run if table has at least 2 rows and 2 columns
  if (nrow(tbl) > 1 && ncol(tbl) > 1) {
    
    test <- suppressWarnings(chisq.test(tbl))
    
    return(data.frame(Feature = col, p_value = test$p.value))
    
  } else {
    
    return(data.frame(Feature = col, p_value = NA))
  }
  
}) %>% bind_rows()

# Sort by significance
chi_results <- chi_results %>% arrange(p_value)

print(chi_results)

sig_cats <- chi_results %>% filter(!is.na(p_value) & p_value < 0.05)

```


#Bar Plots for Significant Categorical Features

```{r,Bar Plots for Significant Categorical Features, echo=TRUE}
# Plot only significant categorical features
for (feature in sig_cats$Feature) {
  
  p <- ggplot(df, aes_string(x = feature, fill = factor(df$readmitted))) +
    
    geom_bar(position = "fill") +
    
    scale_fill_manual(values = c("0" = "skyblue", "1" = "tomato"),
                      
                      labels = c("No", "Within 30 days")) +
    
    labs(title = paste("Readmission Proportions by", feature),
         
         x = feature, y = "Proportion") +
    
    theme_minimal() +
    
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
}
```


## Facet Plot for Multiple Categorical Features

```{r,Facet Plot for Multiple Categorical Features,echo=TRUE}
df %>%
  dplyr::select(all_of(sig_cats$Feature), readmitted) %>%
  
  gather(key = "Feature", value = "Value", -readmitted) %>%
  
  ggplot(aes(x = Value, fill = factor(readmitted))) +
  
  geom_bar(position = "fill") +
  
  facet_wrap(~Feature, scales = "free") +
  
  scale_fill_manual(values = c("0" = "skyblue", "1" = "tomato"),
                    
                    labels = c("No", "Within 30 days")) +
  
  theme_minimal() +
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


#Combine Results for Feature Selection

```{r}
#Combine Results for Feature Selection
list(
  Significant_Categorical = sig_cats$Feature,
  Significant_Numeric = sig_nums$Feature
)
```


#**Dynamic re-coding of all variables to NUMERIC**

```{r}

# Re-code all categorical types (character, factor, ordered factor) to numeric

df_numeric_cats <- df %>%
  
  mutate(across(where(~ is.character(.) | is.factor(.) | is.ordered(.)),~ as.numeric(as.factor(.))))
```


```{r}
#Correlation Matrix

library(corrplot)
df_numeric_cats<-na.omit(df_numeric_cats)

cor_matrix <- round(cor(df_numeric_cats, use = "complete.obs"), digits = 1)


#pdf("correlation_plot.pdf", width = 8, height = 8)

corrplot(cor_matrix, method = "shade", addCoef.col="lightgreen", type="upper", tl.cex = 0.4, number.cex=0.4)

```


```{r}
## Create a flex table with the Correlation Matrix

Cor_Table<-data.frame(cor_matrix)

flextable(Cor_Table)
Cor_Table <- cbind(Feature = rownames(Cor_Table), Cor_Table)
rownames(Cor_Table) <- NULL

ft <- flextable(Cor_Table) %>%
  autofit() %>%
  set_header_labels(Feature = "Feature / Variable")
ft
# Convert flextable back to data frame
df_export <- as.data.frame(ft$body$dataset)
```


#Modelling pipeline (df_sig)

```{r}
df_sig<-df[,c("race","age",  #Demographic Info
              "time_in_hospital","number_inpatient","num_procedures",   #History
              "num_lab_procedures","number_diagnoses","change","A1Cresult", #History
              "number_emergency", "discharge_disposition_id","admission_type_id","number_outpatient" ,  #History
              "num_medications", "max_glu_serum", "repaglinide", "glipizide" , "insulin", "metformin", "diabetesMed", "readmitted")] #medications

# ensure target is factor with levels "no" and "yes"
df_sig <- df_sig%>%
  mutate(readmitted = factor(ifelse(readmitted==1, "yes", "no"), levels = c("no","yes")))

```


#Prepare data & split

```{r,data split, echo=TRUE}

set.seed(22061463)
##stratified train/test data split    
train_data <- df_sig %>%
  
  group_by(readmitted) %>%
  
  
  sample_frac(0.8) %>%
  
  ungroup()

test_data <- anti_join(df_sig, train_data, by = colnames(df_sig))
```


```{r}
#Check proportions**

prop.table(table(train_data$readmitted))

prop.table(table(test_data$readmitted)) 

```



**models**

```{r Logistic regression models, echo=FALSE}

# Logistic regression (glm)
log_model <- glm(readmitted ~ ., 
                 data = train_data, 
                 family = binomial)  
summary(log_model)
```


```{r}
library(dplyr)
library(ranger)
library(pROC)
library(smotefamily)
library(xgboost)
library(caret)
```



```{r}
library(caret)
library(pROC)

evaluate_from_probs <- function(pred_probs, y_true_factor, threshold = 0.3) {

  # ensure proper factor coding
  y_true_factor <- factor(y_true_factor, levels = c("no","yes"))
  pred_class <- factor(ifelse(pred_probs >= threshold, "yes", "no"),
                       levels = c("no","yes"))

  cm <- confusionMatrix(pred_class, y_true_factor, positive = "yes")

  roc_obj <- roc(response = y_true_factor,
                 predictor = pred_probs,
                 levels = c("no","yes"),
                 direction = "<")

  data.frame(
    threshold   = threshold,
    accuracy    = as.numeric(cm$overall["Accuracy"]),
    precision   = as.numeric(cm$byClass["Precision"]),
    recall      = as.numeric(cm$byClass["Recall"]),
    specificity = as.numeric(cm$byClass["Specificity"]),
    f1          = as.numeric(cm$byClass["F1"]),
    auc         = as.numeric(auc(roc_obj))
  )
}
```







## MODEL 1: Logistic Regression (on IMBALANCED data)

```{r}
set.seed(421)
log_model <- glm(readmitted ~ .,data   = train_data,   # <- imbalanced
  family = binomial)

log_pred_probs <- predict(log_model, newdata = test_data, type = "response")

log_results <- evaluate_from_probs(
  pred_probs    = log_pred_probs,
  y_true_factor = test_data$readmitted,
  threshold     = 0.3 #it was changed from 0.5 to 0.3 
)
log_results
```



## 2. SMOTE (Random Forest training)

```{r}
# Prepare numeric version for smotefamily::SMOTE
train_data_numeric <- train_data %>% 
  mutate(readmitted_num = as.numeric(readmitted) - 1) %>% 
  dplyr::select(-readmitted)

train_data_SMOTE_input <- train_data_numeric %>%
  mutate(across(where(is.factor), as.numeric)) 

# Separate features (X) and target (y)
X <- train_data_SMOTE_input %>% dplyr::select(-readmitted_num)
y <- train_data_SMOTE_input$readmitted_num   # numeric 0/1

set.seed(42)
smote_output <- SMOTE(
  X,
  y,
  K = 3,        # 
  dup_size = 0  # balance to majority class
)

train_data_SMOTE_balanced_raw <- smote_output$data

train_data_SMOTE_balanced <- train_data_SMOTE_balanced_raw %>%
  dplyr::rename(readmitted_num = class) %>%
  mutate(
    readmitted = factor(
      ifelse(readmitted_num == 1, "yes", "no"),
      levels = c("no", "yes")
    )
  ) %>%
  dplyr::select(-readmitted_num)

cat("Original train class proportion:\n")
print(prop.table(table(train_data$readmitted)))

cat("SMOTE-balanced train class proportion:\n")
print(prop.table(table(train_data_SMOTE_balanced$readmitted)))
```



```{r}
set.seed(421)
rf_model <- ranger(
  readmitted ~ .,
  data = train_data_SMOTE_balanced,  # <- SMOTE-balanced
  num.trees = 500,
  mtry = 5,
  min.node.size = 10,
  probability = TRUE
)

rf_pred_probs <- predict(rf_model, data = test_data)$predictions[, "yes"]

rf_results <- evaluate_from_probs(
  pred_probs    = rf_pred_probs,
  y_true_factor = test_data$readmitted,
  threshold     = 0.3
)
rf_results

```



##  MODEL 3: XGBoost (on IMBALANCED data with class weights)

```{r}
set.seed(1235)
# Numeric version of TRAIN data for XGBoost
train_xgb <- train_data %>%
  mutate(readmitted_num = ifelse(readmitted == "yes", 1, 0)) %>%
  mutate(across(
    .cols = where(~ is.factor(.) || is.ordered(.)),
    .fns  = ~ as.numeric(as.factor(.))
  ))

# Numeric version of TEST data for XGBoost (same transformation)
test_xgb <- test_data %>%
  mutate(readmitted_num = ifelse(readmitted == "yes", 1, 0)) %>%
  mutate(across(
    .cols = where(~ is.factor(.) || is.ordered(.)),
    .fns  = ~ as.numeric(as.factor(.))
  ))
# Use train levels to define encoding for both
x_train <- model.matrix(readmitted ~ . - 1, data = train_data)
x_test  <- model.matrix(readmitted ~ . - 1, data = test_data)

y_train <- ifelse(train_data$readmitted == "yes", 1, 0)
y_test  <- ifelse(test_data$readmitted == "yes", 1, 0)

x_train <- as.matrix(dplyr::select(train_xgb, -readmitted, -readmitted_num))
x_test  <- as.matrix(dplyr::select(test_xgb,  -readmitted, -readmitted_num))


# Check dimensions match
ncol(x_train); ncol(x_test)

dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dtest  <- xgb.DMatrix(data = x_test,  label = y_test)

# Class weight for imbalance
pos_weight <- sum(y_train == 0) / sum(y_train == 1)



  dtrain <- xgb.DMatrix(x_train, label = y_train)
  dvalid <- xgb.DMatrix(x_test,  label = y_test)
  
  params <- list(
    objective        = "binary:logistic",
    eval_metric      = "auc",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    scale_pos_weight = pos_weight
  )
  
  xgb_model <- xgb.train(
    params = params,
    data   = dtrain,
    nrounds = 500,
    evals = list(val = dvalid),
    early_stopping_rounds = 20,
    verbose = 0
  )
  
  
  
xgb_pred_probs <- predict(xgb_model, dtest)
y_test_factor  <- factor(ifelse(y_test == 1, "yes", "no"),
                         levels = c("no", "yes"))



```

##  Check multiple thresholds for XGB

```{r}
set.seed(1235)
##  Check multiple thresholds for XGB
thresholds <- c(0.5, 0.3, 0.2, 0.1, 0.05)
xgb_threshold_results <- do.call(
  rbind,
  lapply(thresholds, evaluate_from_probs,
         pred_probs    = xgb_pred_probs,
         y_true_factor = y_test_factor)
)
xgb_threshold_results
```


##  0.3 Threshold for XGB

```{r}
set.seed(1236)
# 0.3 thresholds
xgb_results <- evaluate_from_probs(
  pred_probs    = xgb_pred_probs,
  y_true_factor = y_test_factor,
  threshold     = 0.3
)
xgb_results


```



#Generate ROC plot for XGB

```{r}
library(pROC)

roc_xgb <- roc(
  response  = y_test_factor,
  predictor = xgb_pred_probs,
  levels = c("no","yes"),
  direction = "<"
)

```

```{r}
# Generate the ROC Plot with better label 
plot(roc_xgb, 
     main = "ROC Curve: XGBoost (Weighted)", 
     col = "red", 
     lwd = 2, 
     print.auc = TRUE, 
     print.auc.x = 0.3, print.auc.y = 0.5, # Moves the AUC text
     print.thres = 0.3, 
     print.thres.adj = c(1.1, 1.2),       
     print.thres.cex = 0.8)               

# Add a diagonal reference line
abline(a = 0, b = 1, lty = 2, col = "gray")

```


### 6. FINAL COMPARISON TABLE 

```{r}
summary_table <- bind_rows(
  log_results %>% mutate(Model = "Logistic Regression (imbalanced)"),
  rf_results  %>% mutate(Model = "Random Forest (SMOTE-balanced)"),
  xgb_results %>% mutate(Model = "XGBoost (weighted, thr=0.3)")
) %>%
  dplyr::select(Model, threshold, accuracy, precision, recall, specificity, f1, auc)

summary_table
```

